---
title: "Plotting Ontological Terms"
author: "Daniel Greene"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Ontological Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
```{r echo=FALSE}
knitr::opts_chunk$set(dev="svg")
```

`ontologyPlot` lets you to visualise sets of ontological terms as subgraphs of the full ontology. The function provided for doing this is `onto_plot`. Most of the functions in the package use an `ontology_index` object, as in the `ontologyIndex` package, so we start by loading the package, and an example ontology - the Human Phenotype Ontology (HPO). 

```{r}
suppressPackageStartupMessages(library(ontologyIndex))
suppressPackageStartupMessages(library(ontologyPlot))
data(hpo)
```

## Plotting Terms

A set of terms (character vector of term IDs) can be plotted as a subgraph of the ontology by calling `onto_plot`:
```{r}
onto_plot(hpo, terms=c("HP:0001873", "HP:0011877"))
```

To include all ancestral terms, use the `get_ancestors` function:

```{r}
onto_plot(hpo, terms=get_ancestors(hpo, c("HP:0001873", "HP:0011877")))
```

Terms which just link two terms together can be removed to save space, whilst retaining the structure of the relevant part of the ontology - to do this use `remove_links`:

```{r}
onto_plot(hpo, terms=remove_links(hpo, get_ancestors(hpo, c("HP:0001873", "HP:0011877"))))
```

Node attributes can be passed to `onto_plot`. Here we pass in a character vector of term IDs to be used as the labels and a character vector of colours generated with `rainbow`. Note that the length of each vector passed in as an attribute has the same length as `terms`. However, it is still valid to pass in a vector of length 1, for instance `color="blue"`, because in this case `onto_plot` recognises this as a request to set given attribute for all nodes to the passed value.

```{r}
terms <- remove_links(hpo, get_ancestors(hpo, c("HP:0001873", "HP:0011877")))
onto_plot(hpo, terms=terms, label=terms, fillcolor=rainbow(length(terms)))
```

Functions which automatically select graphical parameters depending on other arguments to `onto_plot` (i.e. `ontology`, `frequencies` and `term_sets` - see `?onto_plot` for further details) can also be passed to `onto_plot`. For instance, here the labels are set by passing the function `official_labels`, labelling the nodes by the official names of the corresponding terms.

```{r}
onto_plot(hpo, terms=terms, label=official_labels)
```

Any attributes can be passed to `onto_plot`. Those which are recognised by the package `Rgraphviz` will be used by `plot`. Some of the attributes include size (`width`), shape (`shape`), font sizes (`fontsize`) and border colour (`color`) (see `?onto_plot` for details).

## Frequencies

A `frequencies` argument giving a population frequency of (a subset of) terms in the ontology can be passed to the function `onto_plot`. If this argument is passed, functions which make use of it in determining graphical parameters can be passed in for those parameters, for instance `colour_by_population_frequency`.

```{r}
frequencies <- seq(from=0, to=1, by=1/length(terms))
names(frequencies) <- terms
onto_plot(hpo, terms=terms, frequencies=frequencies, 
	fillcolor=colour_by_population_frequency)
```

## Term sets
Another argument which can be passed to `onto_plot` is `term_sets`, a list of sets of ontological terms (e.g. a set of HPO encoded phenotypes). If passed, the `terms` argument of `onto_plot` defaults to all the terms which help distinguish the cases (see `?remove_uninformative_terms` for more details).

```{r}
hpo_phenotypes <- list(
	A=c("HP:0001382","HP:0004272","HP:0007917","HP:0004912","HP:0001596"),
	B=c("HP:0001382","HP:0004272","HP:0002165","HP:0004800","HP:0004912"),
	C=c("HP:0004800","HP:0001382","HP:0004912","HP:0007917","HP:0008743")
)

onto_plot(hpo, term_sets=hpo_phenotypes, label=label_by_term_set)
```

To further decorate the plot, we could use the frequency of each term in `hpo_phenotypes` them to colour them.

```{r}
onto_plot(
	hpo, 	
	frequencies=get_term_frequencies(hpo, hpo_phenotypes),
	term_sets=hpo_phenotypes,
	label=label_by_term_set,
	fillcolor=colour_by_frequency)
```

## Edge attributes

Edge attributes recognised by `Rgraphviz` can also be supplied by the `edge_attributes` argument.

```{r}
onto_plot(
	hpo, 	
	frequencies=get_term_frequencies(hpo, hpo_phenotypes),
	term_sets=hpo_phenotypes,
	label=label_by_term_set,
	edge_attributes=list(color="red", lty="dashed"))
```

## Plots for publications

To get the highest quality plots and fine-grained control over the node and edge attributes, the best option is to create the `ontological_plot` with `onto_plot` and write it to a file in `dot` format using the `write_dot` function. The file can then be opened using a layout program, for example the graphviz program (http://www.graphviz.org/), to visualise/export as an image. Individual node and edge attributes can then be modified by editing the file.
